RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R

### 2014-03-25 inst/doc/pubs/manuscript.Rmd
updated manuscript.Rmd with @rvosa's edits, regenerate manuscript.md

---
layout: preprint
#layout: review, 11pt
title: "RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Hilmar Lapp
    affiliation: NESCent
  - name: Scott Chamberlain
    affiliation: SimonFraser
  - name: Rutger Vos
    affiliation: NCB
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center, Durham, NC, USA
  - code: SimonFraser
    address: |
      Department of Biology, Simon Fraser University, Burnaby, Canada
  - code: NCB
    address: | 
      NCB Naturalis, Leiden, the Netherlands
abstract: | 
      1. NeXML promises to be the next generation of phylogenetic
      informatics data and metadata exchange format, offering substantial
      improvements in reliability, extensibility, and richness over
      earlier standards such as Nexus.

@@

- Most R users rely on the nexus inherits the limitations of nexus file format

@@

* Why NeXML -- because Nexus standard isn't a standard.  Developers extend it arbitrarily to meet their needs, making it incompatible <!-- with? -->.  For example, Nexus cannot represent horizontal gene transfer, ambiguous use of symbols, cannot be extended reliably, and there is no way to validate that it is proper Nexus.  NeXML addresses all of these problems [Vos _et al._ 2012] with a state-of-the-art data management format.

* Why RNeXML: Huge ecosystem of phylogenetics packages. (1) Read NeXML.
All would benefit by being able to read NeXML, both as a robust standard
free of the pitfalls of Nexus, and more importantly, as a much richer
source of data thanks to its extendable annotation model.  (2) Write
NeXML.  R packages and R users add rich information to the phylogenies
they work with, but need a platform independent data exchange format to
share and archive this. In particular, we illustrate how NeXML can
be used to distribute both phylogenies and character trait data used in 
many popular phylogenetic comparative methods tools in R.  (3) Annotate 
and extend nexml. `RNeXML` provides convenient tools for adding metadata
to NeXML files, including the automated generation of metadata and examples
of how to extend the NeXML standard to represent new data features without
breaking validation or backwards compatability of the file type.

@@

Previously, this was possible using the `treebase` package [@treebase]
for R, which relied on the nexus parser provided in the `ape` package.
Due to inconsistencies present in the Nexus standard [Vos _et al._
2012], a significant number <!-- State explicitly --> of TreeBASE nexus
files could not be read in this way.  Using the more precisely defined
NEXML standard, every valid NeXML file on TreeBASE can now successfully
be parsed to R objects.

@@

Like nexus files, NeXML files can also contain character data.  While
most researchers will be familiar with the use of character matrices in
nexus files as a container for molecular sequence data used to infer
the phylogenetic tree, we find this ideal for character data used in
comparative methods popular in R. Methods to get extract character data
work much like those for the phylogenetic data:

@@

This function uses the `taxize` R library ([Chamberlain 2013]) to check each taxon
label against the NCBI database.  If a unique match is found, a metadata annotation
is added to the taxon providing the NCBI identification number to the taxonomic unit.
If no match is found, the user is warned to check for possible typographic errors in
the taxonomic labels provided.

@@

We see the title appears as the `content` of this `meta` element.  The `id` attribute
provides us with a unique way to refer to this meta element in this document.  The `datatype`
tells us that the content is a text string (as opposed to a numerical or boolean value).  
The `property` is the attribute that most interests us here; it tells us that the content 
is a title.  The prefix `dc` indicates the vocabulary (called a namespace in XML) to which 
the word `title` belongs.  The prefix is defined elsewhere in the NeXML file, as we can see with:

@@

The built-in vocabularies are just the tip of the iceburg of established vocabularies. Here 
we add an annotation from the `skos` namespace which describes the history of where the
data comes from:

@@

<!-- How could this be improved? Should we use something more formal? 
We want this example to be something other users could reasonably
do themselves without expertise in RDFa or XML, but also to be a good
model case that doesn't cut corners.  -->

@@

```{r}
library(phytools)
plotSimmap(phy)
```

@@

Though other NeXML parsers (for instance, for perl or python) have
not been written explicitly to express simmap data, those parsers will
nonetheless be able to successfully parse this file and expose the simmap
data to the user.

>>>>

---
layout: preprint
#layout: review, 11pt
title: "RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Hilmar Lapp
    affiliation: NESCent
  - name: Scott Chamberlain
    affiliation: SimonFraser
  - name: Rutger Vos
    affiliation: NBC
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center, Durham, NC, USA
  - code: SimonFraser
    address: |
      Department of Biology, Simon Fraser University, Burnaby, Canada
  - code: NBC
    address: | 
      NBC Naturalis, Leiden, the Netherlands
abstract: | 
      1. NeXML promises to be the next generation of phylogenetic
      informatics data and metadata exchange format, offering substantial
      improvements in reliability, extensibility, and richness over
      earlier standards such as NEXUS.

@@

- Most R users rely on the NEXUS inherits the limitations of NEXUS file format

@@

* Why NeXML -- because NEXUS standard isn't a standard.  Developers extend
it arbitrarily to meet their needs, yielding mutually incompatible dialects. 
For example, NEXUS cannot represent horizontal gene transfer, ambiguous
use of symbols, cannot be extended reliably, and there is no way to
validate that it is proper NEXUS.  NeXML addresses all of these problems
[Vos _et al._ 2012] with a state-of-the-art data management format.

* Why RNeXML: Huge ecosystem of phylogenetics packages. (1) Read NeXML.
All would benefit by being able to read NeXML, both as a robust standard
free of the pitfalls of NEXUS, and more importantly, as a much richer
source of data thanks to its extendable annotation model.  (2) Write
NeXML.  R packages and R users add rich information to the phylogenies
they work with, but need a platform independent data exchange format to
share and archive this. In particular, we illustrate how NeXML can
be used to distribute both phylogenies and character trait data used in 
many popular phylogenetic comparative methods tools in R.  (3) Annotate 
and extend nexml. `RNeXML` provides convenient tools for adding metadata
to NeXML files, including the automated generation of metadata and examples
of how to extend the NeXML standard to represent new data features without
breaking validation or backwards compatability of the file type.

@@

Previously, this was possible using the `treebase` package [@treebase]
for R, which relied on the NEXUS parser provided in the `ape` package.
Due to inconsistencies present in the NEXUS standard [Vos _et al._
2012], a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Using the more precisely defined
NEXML standard, every valid NeXML file on TreeBASE can now successfully
be parsed to R objects.

@@

Like NEXUS files, NeXML files can also contain character data.  While
most researchers will be familiar with the use of character matrices in
NEXUS files as a container for molecular sequence data used to infer
the phylogenetic tree, we find this ideal for character data used in
comparative methods popular in R. Methods to get extract character data
work much like those for the phylogenetic data:

@@

This function uses the `taxize` R library ([Chamberlain 2013]) to check each taxon
label against the NCBI database.  If a unique match is found, a metadata annotation
is added to the taxon providing the NCBI identification number to the taxonomic unit.
If no match is found, the user is warned to check for possible typographic errors in
the taxonomic labels provided.  <!-- what if multiple NCBI taxon IDs are found, e.g. 'Aotus'? -->

@@

We see the title appears as the `content` of this `meta` element.  The `id` attribute
provides us with a unique way to refer to this meta element in this document.  The `datatype`
tells us that the content is a text string (as opposed to, say, a numerical or boolean value).  
The `property` is the attribute that most interests us here; it tells us that the content 
is a title.  The prefix `dc` indicates the vocabulary (called a namespace in XML) to which 
the word `title` belongs.  The prefix is defined elsewhere in the NeXML file, as we can see with:

@@

The built-in vocabularies are just the tip of the iceberg of established vocabularies. Here 
we add an annotation from the `skos` namespace which describes the history of where the
data comes from:

@@

<!-- How could this be improved? Should we use something more formal? 
We want this example to be something other users could reasonably
do themselves without expertise in RDFa or XML, but also to be a good
model case that doesn't cut corners. @cboettig -->

<!-- I don't see how this could be improved much without forcing people
to fire op protege (let's not!). As it happens, the Dublin Core URI, 
for example, points to an HTML page so even in that case automatic 
reasoners would not be able to consume the vocabulary and draw inferences
about, e.g. identity of predicates (owl:sameAs and such). @rvosa -->

@@

```{r caption="Stochastic character mapping on a phylogeny, as generated by the phytools package after parsing the simmap-extended NeXML"}
library(phytools)
plotSimmap(phy)
```

@@

Though other NeXML parsers (for instance, for Perl or Python) have
not been written explicitly to express simmap data, those parsers will
nonetheless be able to successfully parse this file and expose the simmap
data to the user.
